Compositional Quantitative Resource Analysis of C Programs

### 2014-06-25 paper/main.tex
more rule explanations

Due to \code{break} and \code{return} statements of Clight, there are
different possible ways to exit a block of code.  We also have to keep
track of the resource specifications of functions.  To account for
this in the logic, our quantitative Hoare triples have the form
$$
\Delta; B; R \vdash
\htriple
  { Q }
  { S }
  { Q' } \; .
$$
The triple $\htriple{ Q }{ S }{ Q' }$ consists of a statement $S$ and
two assertions $Q,Q' : \Assn$.  It corresponds to triples in classic
Hoare logic and the intuitive meaning is as follows.  If $S$ is
executed with starting state $\state$, the empty continuation
$\Kstop$, and at least $P(\state)$ resources available then the
evaluation does not run out of resources and there are at least
$Q(\state')$ resources left if the evaluation terminates in state
$\state'$.  The assertion $B : \Assn$ provides the
postcondition for the case in which the code block $S$ is exited by a
\code{break} statement.  So if the execution is terminated in state
$\state'$ with a \code{break} then $B(\state')$ resources are
available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
case in which the code block $S$ is exited by a $\code{return}~x$
statement.  The integer argument of $R$ is the return value.
Finally, the function context of judgements that we write $\Delta$ is
a mapping from function names to specification of the form
$$
  \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v).
$$
The assertion $P_f \, z \, \vec v$ is the precondition of the function
$f$ and the assertion $Q_f \, z \, v$ is its postcondition.  They are
both parameterized by an arbitary logical variable $z$ (which can be a
tuple) that relates the function arguments with the return value.  The
precondition also depends on $\vec v$, the value of all the arguments
at the function invocation.  Similarly, the postcondition depends on
the return value $v$ of the function.  The use of logical variables to
express relations between different states of an execution is a
standard technique of Hoare logic.

@@

In the rule {\sc L:Call}, accounts for the cost usage of
programs. It enforces that enough stack space is available to call the
function $f$ by adding $M(f)$ to the precondition and the
postcondition. The pre- and postconditions are taken from the context
$\Gamma$. This context is extended using the {\sc L:Abstract} rule
described below. The assertions in the context are parametric with
respect to both the function argument value and the return value. This
allows to specify a bound for a function whose recursion depth depends
on an input parameter.  The argument parameter is instantiated by the
call rule using the result of the evaluation of the argument
expression in the current state. Note the symmetry of the rule {\sc
  L:Call}: $M(f)$ is added on both sides to account for the stack
space that becomes available after the call.  This justifies that the
{\sc L:Return} rule does not account for stack-space release.

Finally, we describe the rules which are not syntax directed. The rule
{\sc L:Abstract} allows to make a proof on any statement with an extra
hypothesis in the derivation context provided that we have a proof
that this hypothesis is true. We can see as the quantitative
interpretation of the usual logical Modus Ponens.

There are two weakening rules available in the quantitative Hoare
logic.  The framing rule {\sc L:Frame} is designed to weaken a
statement by stating that if $S$ needs $P$ bytes to run and leaves $Q$
bytes free at its end, then it can very well run with $P + c$ bytes
and return $Q + c$ bytes. It comes very handy when we want to prove
tight bounds using the $\max$ function as demonstrated
in~\pref{fig:example_derivation}. The consequence {\sc L:Conseq} rule
is directly imported from classical Hoare logics except that instead
of using the logical implication $\implies$ we use the quantitative
$\ge$. This rule indeed weakens the statement since it requires more
resource to run the statement and yields less than what has been
proved to be available after its termination.

>>>>

Due to \code{break} and \code{return} statements of Clight, there are
different possible ways to exit a block of code.  We also have to keep
track of the resource specifications of functions.  To account for
this in the logic, our quantitative Hoare triples have the form
$$
\Delta; B; R \vdash
\htriple
  { Q }
  { S }
  { Q' } \; .
$$
The triple $\htriple{ Q }{ S }{ Q' }$ consists of a statement $S$ and
two assertions $Q,Q' : \Assn$.  It corresponds to triples in classic
Hoare logic and the intuitive meaning is as follows.  If $S$ is
executed with starting state $\state$, the empty continuation
$\Kstop$, and at least $P(\state)$ resources available then the
evaluation does not run out of resources and there are at least
$Q(\state')$ resources left if the evaluation terminates in state
$\state'$.  The assertion $B : \Assn$ provides the
postcondition for the case in which the code block $S$ is exited by a
\code{break} statement.  So if the execution is terminated in state
$\state'$ with a \code{break} then $B(\state')$ resources are
available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
case in which the code block $S$ is exited by a $\code{return}~x$
statement.  The integer argument of $R$ is the return value.
Finally, the function context of judgements that we write $\Delta$ is
a mapping from function names to specification of the form
$$
  \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v).
$$
The assertion $P_f \, z \, \vec v$ is the precondition of the function
$f$ and the assertion $Q_f \, z \, v$ is its postcondition.  They are
both parameterized by an arbitary logical variable $z$ (which can be a
tuple) that relates the function arguments with the return value.  The
precondition also depends on $\vec v$, the value of all the arguments
at the function invocation.  Similarly, the postcondition depends on
the return value $v$ of the function.  The use of logical variables to
express relations between different states of an execution is a
standard technique of Hoare logic.
%
To ensure soundness, we require that $P_f$ and $Q_f$ do not depend on
the local variables on the call stack, that is, $\forall z \, \vec v
\, \theta \, \theta' \, \gamma . \, P_f \, z \, \vec v \,
(\theta,\gamma) = P_f \, z \, \vec v \, (\theta',\gamma)$.

For two assertions $P,Q : \Assn$, we write $P \models Q$ to if for all
program state $\state$ $P(\state) \ge Q(\state)$.

@@

In the rule {\sc L:Call}, accounts for the execution cost of both
$M_f$ for function calls and $M_r$ for \code{return} statements.  This
justifies that the {\sc L:Return} rule does not account for resource
consumption.  The pre- and postconditions $P_f$ and $Q_f$ are taken
from the function context $\Delta$.  The assertions in the context are
parametric with respect to both the function argument value and the
return value. This allows to specify a bound for a function whose
resource consumption depends on an input parameter.  The argument
parameter is instantiated by the call rule using the result of the
evaluation of the argument variables in the current state.  Recall
that we require that $P_f$ and $Q_f$ do not depend on the local
variables on the call stack, that is, $\forall z \, \vec v \, \theta
\, \theta' \, \gamma . \, P_f \, z \, \vec v \, (\theta,\gamma) = P_f
\, z \, \vec v \, (\theta',\gamma)$.  To transfer potential that
depends on local variables of the callee from the precondition $P$ to
the postcondition $Q$, we use an assertion $A : \Assn$ that is
independent of global variables, that is, $\forall \theta \, \gamma \,
\gamma' . \, A (\theta,\gamma) = A(\theta,\gamma')$.  It is still
possible to express relations between global and local variables using
logical variables (see the following paragraph).

Finally, we describe the rules which are not syntax directed.  There
are two weakening rules available in the quantitative Hoare logic.
The framing rule {\sc L:Frame} is designed to weaken a statement by
stating that if $S$ needs $P$ resources to run and leaves $Q$
resources available after its execution, then it can very well run
with $P + c$ resources and return $Q + c$ resources.  The consequence
{\sc L:Conseq} rule is directly imported from classical Hoare logic
except that instead of using the logical implication $\Rightarrow$ we
use the quantitative $\models$ that point-wise applies $\ge$.  This
rule indeed weakens the statement since it requires more resource to
run the statement and yields less than what has been proved to be
available after its termination.
